Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-23 paper/ms.tex
minor edits

Another popular method uses the fact that, in the limit of evenly-spaced data
and homoscedastic uncertainties, the covariance matrix is ``Toeplitz''
\citep[for example][]{Dillon:2013}.
There are exact methods for solving Toeplitz matrix equations that scale as
$\mathcal{O}(N\,\log N)$ and methods for computing determinants exactly in
$\mathcal{O}(N^2)$ or approximately in $\mathcal{O}(N\,\log N)$
\citep{Wilson:2014}.
The Toeplitz method is, in some ways, more flexible than \celerite\ because it
can be used with any stationary kernel but it requires uniformly spaced data
and the scaling is worse than \celerite\ so it, in general, be less
efficient when applied to large datasets.

@@

Many other approximate methods for scaling GP inference exist \citep[see, for
example,][and references therein]{Wilson:2015a} and we make no attempt to make
our discussion exhaustive.
The key take away here is that \celerite\ provides and \emph{exact} method for
GP inference for a specific but flexible class of one-dimensional kernel
functions.
The closest cousin is the CARMA method with similar strengths and limitations
but, since there is a simpler relationship between the parameters of a
\celerite\ model and its behavior, \celerite\ can be more easily integrated
with physical or physically interpretable models.

@@

Gaussian Process models have been fruitfully applied to many problems in
astronomical data analysis but the fact that the computational cost scales as
the cube of the number of data points has limited their use to relatively
small datasets.
With the linear scaling of \celerite\, we envision the application of Gaussian
processes may be expanded to much larger datasets.
Despite the restrictive form of the \celerite\ kernel, with a sufficient
number of components it is flexible enough to describe a wide range of
astrophysical variability.
In fact, the relation of the \celerite\ kernel to the damped, driven harmonic
oscillator matches simple models of astrophysical variability, and makes the
parameterization interpretable in terms of resonant frequency, amplitude, and
quality factor.
A major drawback of this method is its quadratic scaling with the number of
terms $J$ but, in many cases, small values of $J$ are sufficient.

Our background is in studying transiting exoplanets, a field which has only
recently begun to adopt full covariance matrices in analyzing the noise in
stellar light curves when detecting or characterizing transiting planets
\citep[for example,][]{Carter:2009, Gibson:2012, Barclay:2015, Evans:2015,
Aigrain:2016, Foreman-Mackey:2016b, Grunblatt:2016, Luger:2016}.
All of these analyses have been limited to small datasets or restrictive
kernel choices.
\celerite\ weakens these requirements by providing a scalable method for
computing the likelihood and a physical motivation for the choice of kernel.
\celerite\ can be used to accurately model stellar oscillations using the
relation to the mixture of stochastically driven, damped simple harmonic
oscillators.
As higher signal-to-noise observations of transiting exoplanet systems are
obtained, the effects of stellar variability will more dramatically impact the
correct inference of planetary transit parameters.
We expect that \celerite\ will be important for transit detection
\citep{Pope:2016, Foreman-Mackey:2016b}, transit timing \citep{Agol:2005,
Holman:2005}, transit spectroscopy \citep{Brown:2001}, Doppler beaming
\citep{Loeb:2003, Zucker:2007}, tidal distortion \citep{Zucker:2007}, phase
functions \citep{Knutson:2007, Zucker:2007}, and more.

@@

In the case of $J$ \celerite\ terms, we can check for negative values of the
PSD by solving for the roots of the power spectrum (see \eqalt{celerite-psd}),
abbreviating with $z = \omega^2$:
\begin{equation}
S(\omega)=  \sum_{j=1}^J \frac{q_j\,z + r_j}{z^2+s_j\,z + t_j} = 0
\end{equation}
where
\begin{eqnarray}
q_j &=& a_j\,c_j-b_j\,d_j\\
r_j &=& (d_j^2+c_j^2)(b_j\,d_j+a_j\,c_j)\\
s_j &=& 2(c_j^2-d_j^2)\\
t_j &=& (c_j^2+d_j^2)^2.
\end{eqnarray}
The denominators of each term are positive, so we can multiply through by
$\prod_j \left(z^2+s_jz + t_j\right)$ to find
\begin{equation}
Q_0(z) = \sum_{j=1}^J (q_j z + r_j)\prod_{k \ne j}\left(z^2+s_kz +
    t_k\right) = 0\quad,
\end{equation}
which is a polynomial with order $2\,(J-1)+1$.
With $J=2$, this yields a cubic equation which may be solved exactly for the
roots.

>>>>

Another popular method uses the fact that, in the limit of evenly-spaced data
and homoscedastic uncertainties, the covariance matrix is ``Toeplitz''
\citep[for example][]{Dillon:2013}.
There are exact methods for solving Toeplitz matrix equations that scale as
$\mathcal{O}(N\,\log N)$ and methods for computing determinants exactly in
$\mathcal{O}(N^2)$ or approximately in $\mathcal{O}(N\,\log N)$
\citep{Wilson:2014}.
The Toeplitz method is, in some ways, more flexible than \celerite\ because it
can be used with any stationary kernel but it requires uniformly spaced data
and the scaling is worse than \celerite\ so it, in general, is less
efficient when applied to large datasets.

@@

Many other approximate methods for scaling GP inference exist \citep[see, for
example,][and references therein]{Wilson:2015a} and we make no attempt to make
our discussion exhaustive.
The key take away here is that \celerite\ provides an \emph{exact} method for
GP inference for a specific but flexible class of one-dimensional kernel
functions.
The closest cousin is the CARMA method with similar strengths and limitations
but, since there is a simpler relationship between the parameters of a
\celerite\ model and its behavior, \celerite\ can be more easily integrated
with physical or physically interpretable models.

@@

Gaussian Process models have been fruitfully applied to many problems in
astronomical data analysis but the fact that the computational cost scales as
the cube of the number of data points has limited their use to relatively
small datasets.
With the linear scaling of \celerite\, we envision the application of Gaussian
processes may be expanded to much larger datasets.
Despite the restrictive form of the \celerite\ kernel, with a sufficient
number of components it is flexible enough to describe a wide range of
astrophysical variability.
In fact, the relation of the \celerite\ kernel to the damped, stochastically-driven harmonic
oscillator matches simple models of astrophysical variability, and makes the
parameterization interpretable in terms of resonant frequency, amplitude, and
quality factor.
A major drawback of this method is its quadratic scaling with the number of
terms $J$ but, in many cases, small values of $J$ are sufficient.

Our background is in studying transiting exoplanets, a field which has only
recently begun to adopt full covariance matrices in analyzing the noise in
stellar light curves when detecting or characterizing transiting planets
\citep[for example,][]{Carter:2009, Gibson:2012, Barclay:2015, Evans:2015,
Aigrain:2016, Foreman-Mackey:2016b, Grunblatt:2016, Luger:2016}.
All of these analyses have been limited to small datasets or restrictive
kernel choices.
\celerite\ weakens these requirements by providing a scalable method for
computing the likelihood and a physical motivation for the choice of kernel.
\celerite\ can be used to accurately model stellar oscillations using the
relation to the mixture of stochastically-driven, damped simple harmonic
oscillators.
As higher signal-to-noise observations of transiting exoplanet systems are
obtained, the effects of stellar variability will more dramatically impact the
correct inference of planetary transit parameters.
We expect that \celerite\ will be important for transit detection
\citep{Pope:2016, Foreman-Mackey:2016b}, transit timing \citep{Agol:2005,
Holman:2005}, transit spectroscopy \citep{Brown:2001}, Doppler beaming
\citep{Loeb:2003, Zucker:2007}, tidal distortion \citep{Zucker:2007}, phase
functions \citep{Knutson:2007, Zucker:2007}, and more.

@@

In the case of $J$ \celerite\ terms, we can check for negative values of the
PSD by solving for the roots of the power spectrum; if there are any real,
positive roots, then the power-spectrum goes negative (or zero), and thus does not
represent a valid kernel. We rewrite the power spectrum, \eqalt{celerite-psd}),
abbreviating with $z = \omega^2$:
\begin{equation}
S(\omega)=  \sum_{j=1}^J \frac{q_j\,z + r_j}{z^2+s_j\,z + t_j} = 0
\end{equation}
where
\begin{eqnarray}
q_j &=& a_j\,c_j-b_j\,d_j\\
r_j &=& (d_j^2+c_j^2)(b_j\,d_j+a_j\,c_j)\\
s_j &=& 2(c_j^2-d_j^2)\\
t_j &=& (c_j^2+d_j^2)^2.
\end{eqnarray}
The denominators of each term are positive, so we can multiply through by
$\prod_j \left(z^2+s_jz + t_j\right)$ to find
\begin{equation}
Q_0(z) = \sum_{j=1}^J (q_j z + r_j)\prod_{k \ne j}\left(z^2+s_kz +
    t_k\right) = 0\quad,
\end{equation}
which is a polynomial with order $2\,(J-1)+1$.
With $J=2$, this yields a cubic equation which may be solved exactly for the
roots.
